Modelling BCG vaccination in the UK

### 2019-12-17 rmds/09-model-fitting.Rmd
major model fitting corrections

This chapter details an approach to fitting a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data were grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+ group). Using age-stratified incidence data, versus aggregated data, allows for more complex trends to be identified.

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that underreporting is more likely than over-reporting. The prior for the reporting noise is based on the observed variation between years.

@@

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight posterior distributions. Ad hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and indicative only.

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
           str_to_sentence() %>% 
           str_replace("-uk", "-UK")
    ) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit.",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CrI)")) %>% 
   set_names(., nm = names(.) %>% str_replace_all("Incidence", "Cases")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB cases for years that the model was fitted to.", 
  caption = "Observed versus predicted overall TB cases for years that the model was fitted to. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

```{r 09-age-strat-incidence, fig.scap = "Observed and predicted TB incidence stratified by model age group", fig.cap = "Observed and predicted annaul TB incidence stratified by model age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. The darker ribbon identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median."}
knitr::include_graphics(file.path(plots_path, "overview-states-1990-age-1.png"))
```

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CrI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>% 
  {.}
```

```{r 09-fig-prior-params, fig.scap = "Prior and posterior distributions for fitted model parameters.", fig.cap = "Prior and posterior distributions for fitted model parameters. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$,$c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$"}
knitr::include_graphics(file.path(plots_path, "prior-posterior-overview-1.png"))
```

@@

```{r 09-fig-sens, fig.scap = "Partial rank correlation coefficients for each parameter fitted too.", fig.cap = "Partial rank correlation coefficients for each parameter fitted too. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$, $c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$", out.width = "250px"}
knitr::include_graphics(file.path(plots_path, "incidence-sensitivity-1.png"))
```

>>>>

This chapter details an approach to fitting a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, the parameters that are fitted, and the parameters that are modelled stochastically. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

The primary data source for the model is the reported, age-stratified, UK born TB notifications from 2000 to 2004 as recorded in the Enhanced TB Surveillance (ETS) system (see Chapter \@ref(data)). 2000 to 2004 are the years for which notifications are stratified by UK birth status with universal school-age BCG vaccination. Data were grouped using the age groups present in the dynamic model (5 year age groups up to 49 and then a group from 50-69 and a 70+ group). Using age-stratified incidence data, versus aggregated data, allows for more complex trends to be identified. Non-UK born TB notifications were extracted from the same source for use as an input to the models force of infection (Chapter \@ref(model-development)).

@@

Where $O$ are the observed notifications, $A$ are the incident cases of disease as forecast by the disease transmission model, $E_{\text{syst}}$ is the systematic reporting error, $E_{\text{noise}}$ is the reporting noise, and $\mathcal{N}$ represents the Gaussian (normal) distribution. The priors for the model are defined in Table \@ref(tab:measurement-model). The prior for systematic reporting error is based on the assumption that underreporting is more likely than over-reporting. The prior for the reporting noise is based on the observed variation between years. This observation model is also used when incorporating non-UK born incidence rates into the models force of infection (Chapter \@ref(model-development)). This allows the uncertain in these observations to be properly accounted for in the incidence estimates produced by the fitted model.

@@

### Stochastic parameters {#stochastic-parameters}

Several key data inputs such as incidence in the non-UK born population, coverage of the BCG vaccination program, births, deaths and the contact rate are not perfectly observed, or recorded, and may vary across time. For this reason these parameters are included into the model developed in the last chapter as noise terms. This means that they are resampled for each timestep and so vary stochastically over time. This results in a model that is semi-stochastic rather than being fully deterministic (Chapter \@ref(introduction)). A semi-stochastic model can be defined as a deterministic model that incoperates stochastic elements but that is still solved as a determinsitc system within a given timestep. It is a modelling approach that has been used prevously in the literature when key parameters are uncertain and potentially time varying.[@Funk2016a] For further details of the stochastic parameters included in the model see Chapter \@ref(model-development).

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvenation steps in all scenarios evaluated. This resulted in spuriously tight posterior distributions. Ad hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. All results presented in the following section are based on posterior distributions produced by the model fitting pipeline using the prior distributions specified in the previous chapter. The results are preliminary and indicative only.

@@

```{r 09-scenarios-table}
tibble(Scenario = names(scenario_dics) %>% str_split("_") %>% 
         map_chr(~paste(., collapse = " ")), 
       DIC = unlist(transpose(scenario_dics)$dic)) %>% 
  arrange(DIC) %>% 
  mutate(Scenario = Scenario %>% 
           str_replace("plus", "and") %>% 
           str_replace("baseline", "Baseline") %>% 
           str_replace("transmission", "Transmission variable in young adults (15-29)") %>% 
           str_replace("mixing", "non-UK born mixing variable in young adults (15-29)") %>% 
           str_to_sentence() %>% 
           str_replace("-uk", "-UK")
    ) %>% 
mutate(DIC = round(DIC, 0)) %>% 
kable(
  caption = "DIC values for each scenario evaluated during model fitting - arranged from best to worst model fit. Using 1000 samples from the posterior distribution of the fitted model for each scenario",
  booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  column_spec(c(1), width = "10cm")
```

@@

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CrI)")) %>% 
   set_names(., nm = names(.) %>% str_replace_all("Incidence", "Cases")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB cases for years that the model was fitted to.", 
  caption = "Observed versus predicted overall TB cases for years that the model was fitted to. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

```{r 09-age-strat-incidence, fig.scap = "Observed and predicted TB incidence stratified by model age group", fig.cap = "Observed and predicted annaul TB incidence stratified by model age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. The darker ribbon identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing."}
knitr::include_graphics(file.path(plots_path, "overview-states-1990-age-1.png"))
```

@@

```{r 09-prior-posterior, warning = FALSE, message = FALSE}
sum_prior_posterior %>% 
  set_names(., nm = . %>% 
              str_replace("or", "or (95\\\\%CrI)")) %>%
  mutate(Parameter = latex_param) %>% 
  select(-latex_param) %>% 
kable(
  caption.short = "Prior and posterior medians for fitted model parameters", 
  caption = "Prior and posterior medians for fitted model parameters. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing.",
  booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>% 
  {.}
```

```{r 09-fig-prior-params, fig.scap = "Prior and posterior distributions for fitted model parameters.", fig.cap = "Prior and posterior distributions for fitted model parameters. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$,$c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing."}
knitr::include_graphics(file.path(plots_path, "prior-posterior-overview-1.png"))
```

@@

```{r 09-fig-sens, fig.scap = "Partial rank correlation coefficients for each parameter fitted too.", fig.cap = "Partial rank correlation coefficients for each parameter fitted too. Parameter names in figure are in their coded form. They can be interpreted as the following parameters:  $\\beta_{\\text{young adult}}$, $c_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{eff}}$, $c^{\\text{hist}}_{\\text{half}}$, $M$, $M_{\\text{young adult}}$, $E_{\\text{syst}}$, $E_{\\text{noise}}$, $\\iota_{\\text{scale}}$, and $\\epsilon^{\\text{older-adult}}_L$. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing.", out.width = "250px"}
knitr::include_graphics(file.path(plots_path, "incidence-sensitivity-1.png"))
```

### 2019-12-17 rmds/10-model-results.Rmd
major model fitting corrections

This chapter details the approach used to extrapolate the findings from the previous chapter beyond the change in BCG policy in 2005 and into the future. It first outlines the scenarios considered, then details the assumptions used to expand the time horizon of the model. Finally the impact of each scenario is explored over multiple time horizons. As discussed in the previous chapter these findings are preliminary in nature, meaning quantitative conclusions cannot be drawn and qualitative conclusions must be appropriately caveated.

@@

The results presented in the following section should be considered preliminary because of the low quality of fit achieved in Chapter \@ref(model-fitting). This means that quantitative estimates are unlikely to be accurate. However, the underlying changes in dynamics caused by vaccination may be used for insight into the likely impact of each scenario and therefore there is still some value in exploring these results.

@@

```{r 10-table-future-incidence}
sum_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CrI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast of overall TB incidence for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast of overall TB incidence for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

@@

```{r 10-age-scenarios-fig, fig.scap = "Forecast of TB incidence for each scenario evaluated from 2005 to 2040, stratified by age group.", fig.cap = "Forecast of TB incidence for each scenario evaluated from 2005 to 2040, stratified by age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. Scenarios are differentiated by colour. The darker ribbon for each colour identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median."}
knitr::include_graphics(file.path(plots_path, "overview-scenarios-future-age-1.png"))
```

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CrI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast of overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast of overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

```{r 10-age-scenarios-deaths-fig, fig.scap = "Forecast of TB mortality for each scenario evaluated from 2005 to 2040, stratified by age group, stratified by age group.", fig.cap = "Forecast of TB mortality for each scenario evaluated from 2005 to 2040, stratified by age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. Scenarios are differentiated by colour. The darker ribbon for each colour identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median."}
knitr::include_graphics(file.path(plots_path, "overview-scenarios-deaths-future-age-1.png"))
```

>>>>

This chapter details the approach used to extrapolate 1000 samples from the posterior distribution of the fitted model from the previous chapter (for the best fitting scenario with variability in both transmission and non-UK born mixing) beyond the change in BCG policy in 2005 and into the future. It first outlines the scenarios considered, then details the assumptions used to expand the time horizon of the model. Finally the impact of each scenario is explored over multiple time horizons. As discussed in the previous chapter these findings are preliminary in nature, meaning quantitative conclusions cannot be drawn and qualitative conclusions must be appropriately caveated.

@@

All results presented in the following section are based on 1000 samples from the posterior distribution of the fitted model from the previous chapter (for the best fitting scenario with variability in both transmission and non-UK born mixing) extended beyond the change in BCG policy in 2005 using the assumptions detailed in the previous section. These results should be considered preliminary because of the low quality of fit achieved in Chapter \@ref(model-fitting). This means that quantitative estimates are unlikely to be accurate. However, the underlying changes in dynamics caused by vaccination may be used for insight into the likely impact of each scenario and therefore there is still some value in exploring these results.

@@

```{r 10-table-future-incidence}
sum_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CrI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast of overall TB incidence for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast of overall TB incidence for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown. (95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

@@

```{r 10-age-scenarios-fig, fig.scap = "Forecast of TB incidence for each scenario evaluated from 2005 to 2040, stratified by age group.", fig.cap = "Forecast of TB incidence for each scenario evaluated from 2005 to 2040, stratified by age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. Scenarios are differentiated by colour. The darker ribbon for each colour identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing."}
knitr::include_graphics(file.path(plots_path, "overview-scenarios-future-age-1.png"))
```

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CrI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast of overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast of overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CrI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

```{r 10-age-scenarios-deaths-fig, fig.scap = "Forecast of TB mortality for each scenario evaluated from 2005 to 2040, stratified by age group, stratified by age group.", fig.cap = "Forecast of TB mortality for each scenario evaluated from 2005 to 2040, stratified by age group (0-11). 0-9 refers to 5 year age groups from 0-4 years old to 45-49 years old. 10 refers to those aged between 50 and 69 and 11 refers to those aged 70+. Scenarios are differentiated by colour. The darker ribbon for each colour identifies the interquartile range, whilst the lighter ribbon indicates the 2.5\\% and 97.5\\% quantiles. The line represents the median. Using 1000 samples from the posterior distribution of the fitted model for the scenario with variability in both transmission and non-UK born mixing."}
knitr::include_graphics(file.path(plots_path, "overview-scenarios-deaths-future-age-1.png"))
```
